---
title: "Cyclistic Bike Share Analysis"
author: "Emanuel Fisha Molla"
date: "`r Sys.Date()`"
output: html_document
 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this analysis, I explore the bike-sharing dataset to identify key trends and insights. This report is based on the data collected from Cyclistic, a bike-sharing program, which will help us understand how riders use the service.

## Problem Statement and Objective

The primary objective of this analysis is to determine how different types of users (e.g., casual vs. annual members) utilize the bike-sharing service. Specifically, we aim to answer the following questions:  

- What is the distribution of rides by type of user?  
- How do ride durations differ between casual and annual members?  
- What time patterns can be observed in the bike usage?  

## Data Frame Overview

The dataset contains information about each ride, including the start and end times, user type, and bike usage details.   
```{r include=FALSE}
# Load libraries
library(tidyverse)
library(lubridate)
```

```{r include=FALSE}
# Load the data
data <- read.csv("../data/processed/all_trips_v2.csv")

# Create a data frame with column names and data types, ensuring no duplication
column_info <- data.frame (Data_Type = sapply(data, function(x) class(x)[1]))

# Display the table
table_info<-knitr::kable(column_info, caption = "Column Names and Data Types")


```
```{r}
table_info
```
  
## Exploratory Data Analysis (EDA)  
Now, let's explore the dataset to understand the key features and patterns.  

#### User Type Distribution
We'll start by analyzing the distribution of users.
```{r include=FALSE}
#The data is cleaned and processed prior  to report.
# Count the number of rides for each user type
user_type_count <- data %>%
  group_by(User_type = member_casual) %>%
  summarise(num_rides = n())
user_type_count<- knitr::kable(user_type_count)
```
```{r}
user_type_count
```
#### Ride Duration Analysis  
Next, we analyze the ride durations for each user type.  

```{r include=FALSE}

# Summary of ride durations by user type
duration_summary<- data %>%
  group_by(User_type=member_casual) %>%
  summarise(Average_ride_length = mean(ride_length))
  #aggregate(data$ride_length ~ data$member_casual, FUN = mean)    # Mean

duration_summary<- knitr::kable(duration_summary)
```
```{r}
duration_summary
```

## Data Visualizations
We now visualize the key findings using graphs.  

#### Distribution of Rides by User Type   

```{r include=FALSE, warning=FALSE}
# Plot the number of rides for each user type for each day of the week
num_rides_viz<- data %>%   
  mutate(weekday = wday(started_at, label = TRUE)) %>%   
  group_by(member_casual, weekday) %>%   
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%   
  arrange(member_casual, weekday) %>%   
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +  
  geom_col(position = "dodge", width = 0.7) + # Adjusted bar width
  scale_fill_manual(values = c("dodgerblue3", "orange2")) +  # Custom colors
  labs(title = "Number of Rides by Rider Type and Weekday", 
       x = "Weekday", y = "Number of Rides", 
       fill = "Rider Type") +
  theme_minimal() +  # Clean theme
  theme(axis.text.x = element_text(angle = 0, hjust = 1.2)) 
```
```{r}
num_rides_viz
```

#### Ride Duration Comparison      
```{r, include=FALSE, warning=FALSE}
ride_dur_viz <- data %>%   
  mutate(weekday = wday(started_at, label = TRUE)) %>%   
  group_by(member_casual, weekday) %>%   
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%   
  arrange(member_casual, weekday) %>%   
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +  
  geom_col(position = "dodge", width = 0.7) +  # Adjusted bar width
  scale_fill_manual(values = c("dodgerblue3", "orange2")) +  # Custom colors
  labs(title = "Average Ride Duration by Rider Type and Weekday", 
       x = "Weekday", y = "Average Duration (minutes)", 
       fill = "Rider Type") +
  theme_minimal() +  # Clean theme
  theme(axis.text.x = element_text(angle = 0, hjust = 1.2))
```

```{r}
ride_dur_viz
```
  
## Conclusion and Recommendations
In this analysis, we observed significant differences between annual members and casual riders:  

* Annual members have a higher number of rides, but their rides tend to be shorter on average.  
* Casual riders tend to take longer rides but are less frequent in their usage.  

### Business Recommendations:  
1. Targeting Casual Riders for Conversion: Casual riders could greatly benefit from an annual membership. Although they tend to take longer trips, they are not using the service as frequently. By converting them to annual members, they would enjoy unlimited rides at no additional cost per ride, which might encourage them to use the service more often, especially for longer trips.  

2. Using Digital Media: Digital campaigns, including social media ads and email promotions, could highlight these benefits, especially targeting those who already show interest in regular bike usage. Promoting the convenience and cost-saving aspects of the annual membership could appeal to casual riders.  

3. Maximizing Member Conversion: By offering promotions, discounts, or trial memberships, Cyclistic can encourage casual riders to become long-term members.  

<!-- Footer with extra space -->
<div style="margin-top: 120px; text-align: center;">
  <footer>
      <p>&copy; 2024 Cyclistic Bike Share. All Rights Reserved.</p>
      <p>Full project available on <a href="https://github.com/Emanuel-DevX/Bike-share">GitHub</a></p>
  </footer>
</div>
  